### 2021-02-09 final_report/include/CE.tex
reviewed all appendices

In \cite{divs}, seven classes of divergences are identified: \textit{thematic}, \textit{promotional}, \textit{demotional}, \textit{structural}, \textit{conflational}, \textit{categorial} and \textit{lexical}. 
However, because the author's proposed way to resolve divergences assumes the availability of more than merely syntactic information\footnote{in particular, of a lexicon of Root Lexical Conceptual Structures specifying, for instance, the logical subject, arguments and modifiers of each verb.} and because, working with UD trees, we do not have access to anything but strictly syntactical and morphological annotations, the very task of identifying these distinctions at this level of granularity becomes, in our position, extremely challenging. 
Consequently, we refer to the simpler classification and less formal definitions proposed in \cite{divs2}, where promotional and demotional divergences are merged in a wider-coverage category of \textit{head-swapping} divergences and where lexical divergences, the hardest to handle in the absence of any kind of semantic information, do not appear at all.
In the following, we give definitions and examples for each of these classes of divergences, specifying to what extent and how\footnote{unless otherwise specified, each divergence can be expressed as a set of simple rules in the form of boolean functions taking the two UD trees candidate for alignment as input.} each of them is handled in the proposed language-agnostic CE module. We do not attempt to cover all possible cases, but rather provide a few examples as a proof of concept.

>>>>

In \cite{divs}, seven classes of divergences are identified: \textit{thematic}, \textit{promotional}, \textit{demotional}, \textit{structural}, \textit{conflational}, \textit{categorial} and \textit{lexical}. 
However, because the author's proposed way to resolve divergences assumes the availability of more than merely syntactic information\footnote{in particular, of a lexicon of Root Lexical Conceptual Structures specifying, for instance, the logical subject, arguments and modifiers of each verb.} and because, working with UD trees, we do not have access to anything but strictly syntactical and morphological annotations, the very task of identifying these distinctions at this level of granularity becomes, in our position, extremely challenging. 
Consequently, we refer to the simpler classification and less formal definitions proposed in \cite{divs2}, where promotional and demotional divergences are merged in a wider-coverage category of \textit{head-swapping} divergences and where lexical divergences, the hardest to handle in the absence of any kind of semantic information, do not appear at all.
In the following, we give definitions and examples for each of these classes of divergences, specifying to what extent and how\footnote{unless otherwise specified, each divergence can be expressed as a set of simple rules in the form of boolean functions taking the pair of UD trees candidate for alignment as input. An example of this can be found in Appendix \ref{cconf}.} each of them is handled in the proposed language-agnostic CE module. We do not attempt to cover all possible cases, but rather provide a few examples as a proof of concept.

### 2021-02-09 final_report/include/backmatter/Appendix_2.tex
reviewed all appendices

\section{Installation}
To compile the CA module, we recommend the Haskell Stack\footnote{\url{https://docs.haskellstack.org/en/stable/README/}}. 
To build the project, clone the above mentioned GitHub repository, move into the corresponding directory and run \texttt{stack build}. 
This will generate a haskell library, named \texttt{concept-alignment}, and the following five executables: \smallskip

@@

The program, however, supports a number of command-line options. Most importantly:
\begin{itemize}
    \item \texttt{--file=FILE} can be used to specify where to write the resulting alignments. Unless otherwise specified (see below), the output consists in two new aligned \texttt{.conllu} files, stored at the chosen location, whose names are prefixed with \texttt{SL} and \texttt{TL} respectively
    \item \texttt{--linearize} specifies that concepts should be in the linearized format described in Section \ref{evalign}, useful for evaluation purposes
    \item \texttt{--maxsize=INT} sets the maximum size of the extracted alignments
    \item \texttt{--all} specifies that the selection step described in Section \ref{select} should be skipped
    \item \texttt{--clauses} enables clause segmentation (cf. Section \ref{clauses})
    \item \texttt{--rest} enables a second pass of alignment ``by exclusion'' (cf. Section \ref{exclusion}) 
    \item \texttt{--pharaoh=FILE} is used to specify a file in pharaoh format (cf. Figure \ref{wordalgn}) to use as backup
\end{itemize}

\subsection{\texttt{propagate-concepts}}
CP requires two, instead of three, \texttt{.conllu} files, one with the concepts to propagate and the following two containing the corpus of annotated sentences where their should be looked for. As a consequence, running it with the default parameters should be done as follows: \smallskip

@@

\subsection{\texttt{evalign}}
The evaluation scrip \texttt{evalign} can be run in three different ``modes'': \smallskip

\begin{enumerate}
    \item single-file mode (\texttt{stack exec -- evalign annotated.txt}): given a pre-annotated file in the linearized format described in Section \ref{evalign}, it prints out basic statistics about precision, recall and amount of reusable alignments.
    \item extraction interactive mode (\texttt{stack exec -- evalign extraction annotated.txt new.txt}): given an annotated and a new, possibly yet-to-annotate file containing linearized alignments, it allows interactive minimal annotation of the latter (if needed) and, on top of printing out the basic statitics, it compares the new alignments to the old ones, telling how many correct and incorrect alignments were lost and/or found
    \item propagation interactive mode (\texttt{stack exec -- evalign propagation annotated.ca new.ca}): similar to extraction interactive mode, excepts that the statistics are CP-specific (percentage of successfully propagated alignments, number of errors introduced by CP etc.)
\end{enumerate} \smallskip

@@

\begin{verbatim}
    stack exec -- GenerateGrammar path_to_extract_grammar 
    paths_to_morphodicts paths_to_aligned_conllu_files
\end{verbatim} \smallskip

Here, \texttt{path\_to\_extract\_grammar} should point to the extraction grammar in \texttt{.pgf}\footnote{Portable Grammar Format, generated by the GF compiler.} format. \texttt{paths\_to\_morphodicts} stands for the path of the morphological dictionaries of the languages involved, also in \texttt{.pgf} format. Similarly, \texttt{paths\_to\_aligned\_conllu\_files} is to be replaced with the files containing the concepts in CoNNL-u format. These files should be, in terms of the languages they are written in, in the same order as the morphological dictionaries.

@@

\section{Configuration: modifying the alignment criteria}
As mentioned in Section \ref{criteria}, modifying the alignment criteria the program makes use of requires little effort. 
All criteria are in fact defined in a separate Haskell module \texttt{Criteria}\footnote{\url{https://github.com/harisont/concept-alignment/blob/master/Criteria.hs}} that exports only the list of criteria to be used. 
In the current implementation, it looks like this: \smallskip

@@

Adding new criteria is also simple, but it requires an understanding of the data type \texttt{Criterion}, defined in module \texttt{ConceptAlignment}:

@@

\begin{itemize}
    \item a boolean function \texttt{func} specifying a rule to decide whether two trees should be aligned. For instance Criterion \ref{udmatch} (UD label matching) is implemented as
    \begin{lstlisting}
        udMatch :: UDTree -> UDTree -> Bool
        (RTree n ts) `udMatch` (RTree m us) = 
            udDEPREL n == udDEPREL m
    \end{lstlisting}
    \item a set of \texttt{Reason} \texttt{reas} to be used for ranking the alignments, shown in the linearized files and used to compute criterion-specific statistics. Reasons, and not alignment rules (\texttt{func}s), are what mirrors criteria as defined in Section \ref{criteria} exactly. A \texttt{func} can define in fact a more specific alignment rule, such as ``type of categorial divergence in which an adjective is replaced by adverb'', while a \texttt{Reason} is in practice a more coarse-grained label associated to potentially many rules
    \item two boolean flags, \texttt{headAlign} and \texttt{strict}, the former specifying whether head alignment should be performed for UD tree pairs matching that criterion, the latter marking the criterion as either strict (i.e. to be also used for ``alignment by exclusion'', cf. Section \ref{exclusion}) or not.
\end{itemize}

>>>>

\section{Installation}
To compile the CA module, we recommend the Haskell Stack\footnote{\url{docs.haskellstack.org}}. 
To build the project, clone the above mentioned GitHub repository, move into the corresponding directory and run \texttt{stack build}. 
This will generate a Haskell library, named \texttt{concept-alignment}, and the following five executables: \smallskip

@@

The program, however, supports a number of command-line options. Most importantly:
\begin{itemize}
    \item \texttt{--file=FILE} can be used to specify where to write the resulting alignments. Unless otherwise specified (see below), the output consists in two new aligned \texttt{.conllu} files, stored at the chosen location, whose names are prefixed with \texttt{SL} and \texttt{TL} respectively
    \item \texttt{--linearize} specifies that concepts should be in the linearized format described in Section \ref{evalign}, useful for evaluation purposes
    \item \texttt{--maxsize=INT} sets the maximum size of the extracted alignments
    \item \texttt{--all} specifies that the selection step described in Section \ref{select} should be skipped
    \item \texttt{--clauses} enables clause segmentation (cf. Section \ref{clauses})
    \item \texttt{--rest} enables a second pass of alignment ``by exclusion'' (cf. Section \ref{exclusion}) 
    \item \texttt{--pharaoh=FILE} is used to specify a file in pharaoh format (cf. Figure \ref{wordalgn}) to use as backup.
\end{itemize}

\subsection{\texttt{propagate-concepts}}
CP requires two, instead of three, \texttt{.conllu} files, one with the concepts to propagate and the following two containing the corpus of annotated sentences where they should be looked for. As a consequence, running it with the default parameters is as follows: \smallskip

@@

\subsection{\texttt{evalign}}
The evaluation script \texttt{evalign} can be run in three different ``modes'': \smallskip

\begin{enumerate}
    \item single-file mode (\texttt{stack exec -- evalign linearized.txt}): \\given a pre-annotated file in the linearized format described in Section \ref{evalign}, it prints out basic statistics about precision, recall and amount of reusable alignments.
    \item extraction mode (\texttt{stack exec -- evalign extraction old.txt new.txt}): \\given an annotated and a new, possibly yet-to-annotate file containing linearized alignments, it allows interactive minimal annotation of the latter (if needed) and, on top of printing out the basic statistics, it compares the new alignments to the old ones, telling how many correct and incorrect alignments were lost and/or found
    \item propagation mode (\texttt{stack exec -- evalign propagation new.txt new.txt}): similar to extraction interactive mode, excepts that the statistics are CP-specific (percentage of successfully propagated alignments, number of errors introduced by CP etc.).
\end{enumerate} \smallskip

@@

\begin{verbatim}
    stack exec -- generate-grammar path_to_extract_grammar 
    paths_to_morphodicts paths_to_aligned_conllu_files
\end{verbatim} \smallskip

Here, \texttt{path\_to\_extract\_grammar} should point to the extraction grammar in \texttt{.pgf}\footnote{Portable Grammar Format, generated by the GF compiler.} format. \texttt{paths\_to\_morphodicts} stands for the path of the morphological dictionaries of the languages involved, also in \texttt{.pgf} format.

Similarly, \texttt{paths\_to\_aligned\_conllu\_files} is to be replaced with the files containing the concepts in CoNNL-u format. These files should be, in terms of the languages they are written in, in the same order as the morphological dictionaries.

@@

\section{Configuration: modifying the alignment \\criteria} \label{cconf}
As mentioned in Section \ref{criteria}, modifying the criteria CA makes use of requires little effort. 
All criteria are in fact defined in a separate Haskell module \texttt{Criteria}\footnote{\url{https://github.com/harisont/concept-alignment/blob/master/Criteria.hs}} that exports the list of criteria to be used only. 
In the current implementation, it looks like this: \smallskip

@@

Adding new criteria is simple too, but it also requires an understanding of the data type \texttt{Criterion}, defined in module \texttt{ConceptAlignment}:

@@

\begin{itemize}
    \item a boolean function \texttt{func} specifying a rule to decide whether two trees should be aligned. For instance, Criterion \ref{udmatch} (UD label matching) is implemented as
    \begin{lstlisting}
        udMatch :: UDTree -> UDTree -> Bool
        (RTree n ts) `udMatch` (RTree m us) = 
            udDEPREL n == udDEPREL m
    \end{lstlisting}
    \item a set of ``reasons'' \texttt{reas} to be used for ranking the alignments, shown in the linearized files and used to compute criterion-specific statistics. \texttt{Reason}s, and not alignment rules (\texttt{func}s), are what mirrors alignment criteria (as they are defined in Section \ref{criteria}) exactly. A \texttt{func} can define in fact a more specific alignment rule, such as ``type of categorial divergence in which an adjective is replaced by adverb'', while a \texttt{Reason} is in practice a more coarse-grained label associated to potentially many rules
    \item two boolean flags, \texttt{headAlign} and \texttt{strict}, the former specifying whether head alignment (cf. Section \ref{heads}) should be performed for UD tree pairs matching that criterion, the latter marking the criterion as either strict (i.e. to be also used for ``alignment by exclusion'', cf. Section \ref{exclusion}) or not.
\end{itemize}
